Practical Optional Types for Clojure

### 2015-07-10 formal-model.tex
tweak 3.1

Now that we have demonstrated the core features Typed Clojure
provides, we link them together in a formal model called
\lambdatc{}.

Building on occurrence typing,
we incrementally add each
novel feature of Typed Clojure to the formalism,
interleaving presentation of syntax, typing rules, operational semantics,
and subtyping.

@@

Our presentation will start with a review of
occurrence typing~\cite{TF10}, the foundation of \lambdatc{}.
We build up the occurrence typing calculus for illustrative purposes, 
and present the full syntax at the end of the section.

\paragraph{Logical formulas over type environments}
Standard type systems use type environments to track lexical type information.
Introducing a lexically scoped variable $x$ is a matter of
associating $x$ with its type $\Number$ (a number) in the type environment.
Correspondingly, checking the type of a variable is a simple map lookup.
$$
\begin{array}{lr}
\judgementtwo{\x{} \mapsto \Number}{\appexp{\inc{}}{\x{}}}{\Number}
&
\infer [LC-Local]
{ {\propenv{}}(\x{}) = {\t{}}
}
{ \judgementtwo {\propenv{}} 
            {\x{}} {\t{}}
           }
\end{array}
$$

Occurrence typing takes a different approach.
The same information is conveyed by \emph{logical formulas}.
The logical statement {\isprop{\Number{}}{\x{}}} says the value
obtained via $x$ is of type \Number{}. A \emph{logical system}
must now \emph{prove}
a variable's type.
$$
\begin{array}{lr}
\judgementtwo{\isprop{\Number{}}{\x{}}}{\appexp{\inc{}}{\x{}}}{\Number}
&
\infer [T-LocalProve]
{ \inpropenv {\propenv{}} {\isprop {\t{}} {\x{}}}}
{ \judgementtwo {\propenv{}} 
            {\x{}} {\t{}}
           }
\end{array}
$$
The proof system judgment { \inpropenv {\propenv{}} {\isprop {\t{}} {\x{}}}}
says under logical assumptions {\propenv{}}, variable \x{} is of type \t{}.
%We later define the more general T-Local.

\paragraph{Symbolic representation}
As we saw in \secref{sec:overview},
occurrence typing is capable of reasoning
about deeply nested expressions.
For each expression, the type system assigns a symbolic representation
with respect to the current lexical environment. For example,
\begin{itemize}
  \item variable \makelocal{m} is represented as \makelocal{m}; 
  \item $\appexpone{\ccclass{\appexp{\makekw{lunch}}{\makelocal{m}}}}$ is represented as ${\path{\classpe{}}{\path{\keype{\makekw{lunch}}}{\makelocal{m}}}}$;
  \item $42$ has no representation, called \emptyobject{}.
\end{itemize}
This symbolic representation is called the expression's \emph{object}. 
\figref{main:figure:termsyntax} gives the syntax for objects.
\emptyobject{} is the
\emph{empty object}. Non-empty objects \path{\pathelem{}}{\x{}} consist of a root variable \x{} and a \emph{path} \pathelem{},
a possibly-empty sequence of \emph{path elements} applied right-to-left starting from the root variable.
We use two path elements---{\classpe{}} and {\keype{k}}---representing the result
of a calling \classconst{} and the result of looking up a keyword $k$ respectively.

@@

\paragraph{Logical statements about objects}
So far we have seen logical statements about the simplest object, a variable---{\isprop{\t{}}{\x{}}}---
saying variable \x{} is of type \t{}.
We extend logical statements to reason about \emph{arbitrary} non-empty 
objects {\isprop{\t{}}{\path{\pathelem{}}{\x{}}}}.

@@

\paragraph{Logical system in action} 
To attain the type of \makelocal{o}, 
T-LocalProve says we must solve for \t{} in
$
{ \inpropenv 
  {\propenv{}}
  {\isprop {\t{}} {\makelocal{o}}}}
$,
under proposition environment 
$$
\propenv{} = {{\isprop{\Order}{\makelocal{o}}},
    {\isprop{\Value{\makekw{combo}}}{\path{\keype{\makekw{Meal}}}{\makelocal{o}}}}}
$$

The main 
$$
\LUpdatePos
$$
Now the purpose of a logical representation can be conveyed---we rely on the logical
system to combine pieces of type information to deduce more accurate types for lexical
bindings.

@@

\paragraph{`Then' and `else' propositions} Along with an object, each expression carries two
propositions---a `then' and an `else' proposition,
that can be assumed if the expression evaluates to a
true value and a false value respectively.

The `then' proposition {\thenprop{\prop{}}} for the variable $o$ is $\notprop{\falsy{}}{o}$, for example.
This says if the expression $o$ evaluates to a true value, then $o$ is neither
of type \False or \Nil. Naturally, the `else' proposition {\elseprop{\prop{}}} is $\isprop{\falsy{}}{o}$---which
is true when $o$ is a false value---says $o$ is either \False or \Nil.

@@

\paragraph{Expressions} The syntax is given in \figref{main:figure:termsyntax}. Expressions include variables, values,
application, abstractions, conditionals, and let expressions.
All binding forms introduce fresh variables.
Values include booleans, \nil{}, class literals, keywords, 
numbers,
constants, and strings. Lexical closures {\closure {\openv{}} {\abs {\x{}} {\t{}} {\e{}}}}
close over functions with value environments \openv{}, which map bindings to values.

@@

The branches are checked with T-False and T-True respectively,
%$$
%\judgementselfrewrite{\propenv{},{\isprop{\Number{}}{\makelocal{d}}}}
%  {\false{}}
%  {\False{}}
%  {\filterset{\botprop{}}
%             {\topprop{}}}
%  {\emptyobject{}}
%$$
then T-Subsume.
$$
\begin{array}{c}
\judgementselfrewrite{\propenv{},{\isprop{\Number{}}{\makelocal{d}}}}
  {\false{}}
  {\Boolean{}}
  {\filterset{\notprop{\Number{}}{\makelocal{d}}}
             {\isprop{\Number{}}{\makelocal{d}}}}
  {\emptyobject{}}
  \\
\judgementselfrewrite{\propenv{},{\notprop{\Number{}}{\makelocal{d}}}}
  {\true{}}
  {\Boolean{}}
  {\filterset{\notprop{\Number{}}{\makelocal{d}}}
             {\isprop{\Number{}}{\makelocal{d}}}}
  {\emptyobject{}}
\end{array}
$$
%to suit the T-If outputs \t{} = \Boolean{}, \thenprop{\prop{}}
%= {\notprop{\Number{}}{\makelocal{d}}}, \elseprop{\prop{}} = {\isprop{\Number{}}{\makelocal{d}}},
%and \object{} = {\emptyobject{}}.
%
%In T-Subsume, we can upcast \t{} = \False{} to \tp{} = \Boolean{} via the premise 
%\issubtypein{}{\t{}}{\tp{}}.
%and 
%\inpropenv {\propenv{}, {\elseprop {\prop{}}}} {\elseprop {\propp{}}}
%from 
%{\elseprop {\prop{}}} = \topprop{} to {\elseprop {\propp{}}} = {\isprop{\Number{}}{\makelocal{d}}}.
The T-Subsume conditions
\inpropenv {\propenv{}, {\thenprop {\prop{}}}} {\thenprop {\propp{}}}
and
\inpropenv {\propenv{}, {\elseprop {\prop{}}}} {\elseprop {\propp{}}}
allow us to pick compatible propositions.

@@

\subsection{Proof system}

>>>>

Now that we have demonstrated the core features Typed Clojure
provides, we link them together in a formal model called
\lambdatc{}.
%
Building on occurrence typing,
we incrementally add each
novel feature of Typed Clojure to the formalism,
interleaving presentation of syntax, typing rules, operational semantics,
and subtyping.

@@

Our presentation will start with a review of
occurrence typing~\cite{TF10}, the foundation of \lambdatc{}.
%We build up the occurrence typing calculus for illustrative purposes, 
%and present the full syntax at the end of the section.

\paragraph{Logical formulas over type environments}
In standard type systems, the 
%use type environments to track lexical type information.
rules for variables use \emph{type environments}.
%associating $x$ with its type $\Number$ (a number) in the type environment,
%%Correspondingly, 
%and checking the type of a variable is a simple map lookup.
$$
\begin{array}{lr}
\infer [LC-Let]
{ \judgementtwo {{\propenv{}}}
                {\e{1}} {\s{}}
  \\
  \judgementtwo {{\propenv{}},\x{} \mapsto {\s{}}}
                {\e{2}} {\t{}}
}
{ 
  \judgementtwo {\propenv{}} 
            {\letexp{\x{}}{\e{1}}{\e{2}}} {\t{}}
           }
&
\infer [LC-Local]
{ {\propenv{}}(\x{}) = {\t{}}
}
{ \judgementtwo {\propenv{}} 
            {\x{}} {\t{}}
           }
\end{array}
$$

Occurrence typing instead uses \emph{logical formulas}
and a \emph{proof system}.
The logical statement {\isprop{\s{}}{\x{}}} says
variable $x$ is of type \s{}. 
%A \emph{logical system}
%must now \emph{prove}
%a variable's type.
$$
\begin{array}{lr}
\infer [T0-Let]
{ \judgementtwo {{\propenv{}}}
                {\e{1}} {\s{}}
  \\
  \judgementtwo {{\propenv{}},\isprop{\s{}}{\x{}}}
                {\e{2}} {\t{}}
}
{ 
  \judgementtwo {\propenv{}} 
            {\letexp{\x{}}{\e{1}}{\e{2}}}
            {\replacefor{\t{}}{\emptyobject{}}{\x{}}}
           }
%\judgementtwo{\isprop{\Number{}}{\x{}}}{\appexp{\inc{}}{\x{}}}{\Number}
&
\infer [T0-Local]
{ \inpropenv {\propenv{}} {\isprop {\t{}} {\x{}}}}
{ \judgementtwo {\propenv{}} 
            {\x{}} {\t{}}
           }
\end{array}
$$
The proof system judgment { \inpropenv {\propenv{}} {\isprop {\t{}} {\x{}}}}
says under logical assumptions {\propenv{}}, variable \x{} is of type \t{}.
Types can now refer to variables, so we erase \x{} from the final type
in T0-Let.
%We later define the more general T-Local.

\paragraph{Symbolic representation}
%As we saw in \secref{sec:overview},
%occurrence typing is capable of reasoning
%about deeply nested expressions.
To reason about nested expressions,
each expression is associated with 
a symbolic representation
called an \emph{object}.
%with respect to the current lexical environment. 
For example,
  variable \makelocal{m} has object \makelocal{m};
  $\appexpone{\ccclass{\appexp{\makekw{lunch}}{\makelocal{m}}}}$ has object ${\path{\classpe{}}{\path{\keype{\makekw{lunch}}}{\makelocal{m}}}}$; and
  $42$ has the \emph{empty} object \emptyobject{}.

\figref{main:figure:termsyntax} gives the syntax for objects---non-empty objects 
\path{\pathelem{}}{\x{}} consist of a root variable \x{} and a \emph{path} \pathelem{},
consisting of
a possibly-empty sequence of \emph{path elements} applied right-to-left starting from the root variable.
We use two path elements---{\classpe{}} and {\keype{k}}---representing the result
of a calling \classconst{} and the result of looking up a keyword $k$ respectively.

@@

\paragraph{Logical statements about objects}
%So far we have seen logical statements about the simplest object, a variable---{\isprop{\t{}}{\x{}}}---
%saying variable \x{} is of type \t{}.
We extend logical statements to reason about \emph{arbitrary} non-empty 
objects {\isprop{\t{}}{\path{\pathelem{}}{\x{}}}}.

@@

\paragraph{Logical system in action} 
To attain the type of \makelocal{o}, 
T0-Local says we must solve for \t{} in
$
{ \inpropenv 
  {\propenv{}}
  {\isprop {\t{}} {\makelocal{o}}}}
$,
under proposition environment 
$$
\propenv{} = {{\isprop{\Order}{\makelocal{o}}},
    {\isprop{\Value{\makekw{combo}}}{\path{\keype{\makekw{Meal}}}{\makelocal{o}}}}}
$$
which deduces \t{} to be a {\makekw{combo}} meal.
The logical
system \emph{combines} pieces of type information to deduce more accurate types for lexical
bindings---this is explained in \secref{formalmodel:proofsystem}.

@@

\paragraph{Expressions} The syntax is given in \figref{main:figure:termsyntax}. Expressions include variables, values,
application, abstractions, conditionals, and let expressions.
All binding forms introduce fresh variables.
Values include booleans, \nil{}, class literals, keywords, 
numbers,
constants, and strings. Lexical closures {\closure {\openv{}} {\abs {\x{}} {\t{}} {\e{}}}}
close value environments \openv{} over functions, which map bindings to values.

\paragraph{Branching propositions} 
%Expressions in test position communicate propositions to each branch.
Each test expression \e{1} in
\ifexp{\e{1}}{\e{2}}{\e{3}} has a `then' and an `else' proposition to check
\e{2} and \e{3} respectively.
For example, in \ifexp{\makelocal{o}}{\e{2}}{\e{3}}
we know {\makelocal{o}} is true in \e{2} and false in \e{3}.
%that can be assumed if the expression evaluates to a
%true value and a false value respectively.
%
The `then' and `else' propositions 
for \makelocal{o}
are $\notprop{\falsy{}}{\makelocal{o}}$ and $\isprop{\falsy{}}{\makelocal{o}}$ respectively.

@@

@@

The branches are checked with T-False and T-True respectively,
%$$
%\judgementselfrewrite{\propenv{},{\isprop{\Number{}}{\makelocal{d}}}}
%  {\false{}}
%  {\False{}}
%  {\filterset{\botprop{}}
%             {\topprop{}}}
%  {\emptyobject{}}
%$$
then T-Subsume.
$$
\begin{array}{c}
\judgementselfrewrite{\propenv{},{\isprop{\Number{}}{\makelocal{d}}}}
  {\false{}}
  {\Boolean{}}
  {\filterset{\notprop{\Number{}}{\makelocal{d}}}
             {\isprop{\Number{}}{\makelocal{d}}}}
  {\emptyobject{}}
  \\
\judgementselfrewrite{\propenv{},{\notprop{\Number{}}{\makelocal{d}}}}
  {\true{}}
  {\Boolean{}}
  {\filterset{\notprop{\Number{}}{\makelocal{d}}}
             {\isprop{\Number{}}{\makelocal{d}}}}
  {\emptyobject{}}
\end{array}
$$
%to suit the T-If outputs \t{} = \Boolean{}, \thenprop{\prop{}}
%= {\notprop{\Number{}}{\makelocal{d}}}, \elseprop{\prop{}} = {\isprop{\Number{}}{\makelocal{d}}},
%and \object{} = {\emptyobject{}}.
%
%In T-Subsume, we can upcast \t{} = \False{} to \tp{} = \Boolean{} via the premise 
%\issubtypein{}{\t{}}{\tp{}}.
%and 
%\inpropenv {\propenv{}, {\elseprop {\prop{}}}} {\elseprop {\propp{}}}
%from 
%{\elseprop {\prop{}}} = \topprop{} to {\elseprop {\propp{}}} = {\isprop{\Number{}}{\makelocal{d}}}.
The T-Subsume premises
\inpropenv {\propenv{}, {\thenprop {\prop{}}}} {\thenprop {\propp{}}}
and
\inpropenv {\propenv{}, {\elseprop {\prop{}}}} {\elseprop {\propp{}}}
allow us to pick compatible propositions for both branches.

@@

\label{formalmodel:proofsystem}
\subsection{Proof system}
